\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-18 paper/algorithm.tex
paper: clean up some TODOs for cost model

\subsection{Cost model}
\label{sec:cost-model}
\todo{Explain that loss is now averaged}
\todo{Do we want to keep the behavior that imputation increases in quality with the number of cells, or switch to number of tuples.}
We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \textsc{Loss}(q).\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance. $\alpha = 1.0$ means that the optimal query should have the highest quality imputation (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast as possible (at the expense of imputation quality).

@@

\todobox{this section must be rewritten to show new approach}{
For each sub-query, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

\begin{align*}
  \textsc{Loss}(q) = \begin{cases}
    \sum_{c \in C} \textsc{Missing}(c, q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{|q'|}} \sum_{c \in C} \textsc{Missing}(c, q') & q = \mu_C(q') \\
    \textsc{Loss}(q_1') + \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.
}

To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
database for the time. Scan costs are estimated based on the number of tuples, 
page size, and an unit-less IO cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

\todobox{This seems tailored to decision trees, but we should talk more generally since
we claim the imputation step isn't the focus}{
Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
different decision tree building algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans), though the range of $\alpha$ that does promote tradeoffs is condensed.
}

>>>>

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \frac{\Sigma_i \textsc{Loss}(q)_i}{|\textsc{Loss}(q)|}.\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(q)$ is an estimate of the amount of error introduced by a sequence of imputation procedure. Our
cost model averages over the losses from each such operator. 
$\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance. $\alpha = 1.0$ means that the optimal query should have the highest quality imputation (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast as possible (at the expense of imputation quality).

@@

For each sub-query, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute the set $\text{Loss}(Q)$ as follows.

\begin{align*}
  \textsc{Loss}(q) = \begin{cases}
    \{1\} \cup \textsc{Loss}(q') & q = \delta_C(q') \\
    \{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Cardinality}(q') }} \} \cup \textsc{Loss}(q')  & q = \mu_C(q') \\
    \textsc{Loss}(q_1') \cup \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
  \end{cases}
\end{align*}

Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
database. Scan costs are estimated based on the number of tuples, 
page size, and an unit-less IO cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the base implementation of our system,
we use an iterative algorithm (see Sec.~\ref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

In order to avoid having the time component of our cost model dominate the information loss 
component, we scale the former by an arbitrary constant.

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).
